---
title: "Guerry data: Maps with sf and ggplot2"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: Guerry
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Guerry data: Maps with sf and ggplot2}
  %\VignetteKeywords{maps, sf, ggplot2, choropleth, France}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi = 96
)
```

The maps in this package (`gfrance`, `gfrance85`) were originally built as `sp`
`SpatialPolygonsDataFrame` objects, and the other vignettes and the README use the
older `sp::spplot()` function to display them. The current standard toolchain for
spatial visualization in R is the [`sf`](https://r-spatial.github.io/sf/) package,
together with `ggplot2::geom_sf()`. This short vignette shows how to work with
Guerry's map data that way.

This doesn't replace the `sp`/`spplot()` examples used elsewhere in the package --
both remain fully supported -- it's simply the modern alternative for anyone
building on `Guerry`'s map data in a `ggplot2` workflow.

```{r load}
library(Guerry)
library(sf)
library(ggplot2)
library(dplyr)
library(tidyr)
data(gfrance85)
data(Guerry_ranks)
```

## Converting to `sf`

Converting the shipped `sp` object to `sf` is a single call to `sf::st_as_sf()`:

```{r st_as_sf}
gf_sf <- st_as_sf(gfrance85)
class(gf_sf)
```

The result is an ordinary data frame with a `geometry` list-column, so it can be
manipulated with usual tools (`dplyr`, `tidyr`, ...) before plotting.

## A basic map

`geom_sf()` handles the geometry column automatically; no need to supply `x`/`y`
aesthetics.

```{r sf-basic}
ggplot(gf_sf) +
  geom_sf(fill = "grey90", color = "white") +
  theme_void()
```

## Choropleth maps

Mapping a variable to `fill` gives a choropleth map. Here, `Literacy` (percent of
military conscripts who could read and write), using the same `PuBu` palette used
elsewhere in this package for "more is better" variables.

```{r sf-choropleth}
ggplot(gf_sf) +
  geom_sf(aes(fill = Literacy), color = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "PuBu", direction = 1, name = "Literacy") +
  theme_void()
```

## Small multiples of the main variables

The package README shows a static image of six choropleth maps of Guerry's main
"moral variables". Here is a living, code-generated version of the same idea,
using `facet_wrap()` and the pre-ranked `Guerry_ranks` data set (rank 1 = best,
86 = worst on each variable):

```{r sf-facet, fig.width=8, fig.height=6}
main_vars <- c("Crime_pers", "Crime_prop", "Literacy", "Donations", "Infants", "Suicides")

ranks_sf <- gf_sf |>
  select(dept) |>
  left_join(Guerry_ranks |> select(dept, all_of(main_vars)), by = "dept") |>
  pivot_longer(cols = all_of(main_vars), names_to = "variable", values_to = "rank")

ggplot(ranks_sf) +
  geom_sf(aes(fill = rank), color = NA) +
  facet_wrap(~ variable) +
  scale_fill_distiller(palette = "PuBu", direction = -1, name = "Rank") +
  theme_void() +
  theme(strip.text = element_text(size = 11, face = "bold"))
```

## Region map with department labels

Finally, a map colored by `Region`, with department names added via
`geom_sf_text()` -- the `sf`/`ggplot2` equivalent of the `plot()` + `text()`
approach used for this same map in the README.

```{r sf-region, fig.width=7, fig.height=7}
col.region <- colors()[c(149, 254, 468, 552, 26)]  # same colors used in the README

ggplot(gf_sf) +
  geom_sf(aes(fill = Region), color = "white", linewidth = 0.3) +
  geom_sf_text(aes(label = Department, color = Region == "W"),
               size = 2.2, check_overlap = TRUE) +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  scale_fill_manual(values = col.region) +
  theme_void()
```

## Next steps

A themed, styled version of these maps using the historical color palettes and
patterns of the
[`ggCheysson`](https://github.com/friendly/ggCheysson) package is planned once
that package reaches CRAN.
