---
title: "Guerry data: Multivariate Analysis"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Guerry data: Multivariate Analysis}
  %\VignetteKeywords{crime, literacy, suicide, France, multivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,   # suppress package loading messages
  comment = "#>",
  fig.height = 5,
  fig.width = 5
)
```

André-Michel Guerry's _Essai sur la Statistique Morale de la France_ [@Guerry:1833]
collected data on crimes, suicide, literacy and other "moral statistics" for various départements
in France. He provided the first real social data analysis, using graphics and maps to summarize this
multivariate dataset. One of his main goals in this ground-breaking study was to determine if the prevalence
of crime in France could be explained by other social variables.

In 1833, the scatterplot had not yet been invented; the idea of a correlation or a regression was still 50 years
in the future [@Galton:1886]. Guerry displayed his data in shaded choropleth maps and semi-graphic tables
and argued how these could be seen as implying systematic, lawful relations among moral variables.

In this analysis, we ignore the spatial context of the départements and focus on multivariate analyses of the
the data set.

## Load the data and packages

```{r load}
library(Guerry)
library(car)
library(ggplot2)
library(ggrepel)
data(Guerry)
```

## `Guerry` data set

Guerry's (1833) data consisted of six main moral variables shown in the table below. 
He wanted all of these to be recorded on aligned scales so that **larger** numbers consistently
reflected "**morally better**". Thus, four of the variables are recorded in the inverse form,
"Population per ...".


| Name        | Description |
|:------------|:------------|
|`Crime_pers` | Population per crime against persons|
|`Crime_prop` | Population per crime against property|
|`Literacy`   | Percent of military conscripts who can read and write|
|`Donations`  | Donations to the poor|
|`Infants`    | Population per illegitimate birth|
|`Suicides`   | Population per suicide|

The `Guerry` data set also contains:

* `dept` and `Department`, the French ID numbers and names for the 86 départements of metropolitan France in 1830, including Corsica.
* `Region`: a factor with main levels "N", "S", "E", "W", "C". Corsica is coded as `NA`.
* A collection of 14 other related variables from other sources at the same time. See `?Guerry` for their precise definitions.

```{r guerry-more-vars}
names(Guerry)[-(1:9)]
```

Among these, as other aspects of criminal behavior,  we see crime against parents,  `infanticide` and `prostitutes`. 

## Guerry's questions

The main questions that concerned Guerry were whether indicators of crime could be shown to be related to factors
which might be considered to ameliorate crime. Among these, Guerry focused most on `Literacy` defined as the
number of military conscripts who could do more than mark an "X" on their enrollment form.
A related variable is `Instruction`, the rank recorded from Guerry's map;
as defined, it is inversely related to `Literacy`.

Other potential explanatory variables are:

: `Donations` (a measure of donations to the poor),

: `Clergy` (the rank of number of Catholic priests in active service, per population)


## Multivariate visualization methods

Visualization methods for multivariate data take an enormous variety of forms simply because more than two dimensions of data offer exponentially increasingly possibilities.  It is useful to distinguish several broad categories:

* __data plots__ : primarily plot the raw data, often with annotations to aid interpretation (regression lines and smooths, data ellipses, marginal distributions)

* __model plots__ : primarily plot the results of a fitted model, considering that
the fitted model may involve more variables than can be shown in a static 2D plot.
Some examples are: Added variable plots, ...

* __diagnostic plots__ : indicating potential problems with the fitted model.

* __dimension reduction plots__ : plot representations of the data into a space of fewer dimensions than the number of variables in the data set. Simple examples include principal components analysis (PCA) and the related biplots,
and multidimensional scaling (MDS) methods.

## Data plots

Data plots portray the data in a space where the coordinate axes are the observed variables.

### Bivariate relationships

Let's start with plots of crime (`Crime_pers` and `Crime_prop`) in relation to `Literacy`.
A simple scatterplot is not very informative. All that can be seen is that there is not much of a relation between personal crime and literacy.

```{r lit-pers-scat0}
ggplot(aes(x=Literacy, y=Crime_pers/1000), data=Guerry) +
  geom_point(size=2) 

```


More useful scatterplots are annotated with additional
statistical summaries to aid interpretation:

* linear regression line,
* smoothed non-parametric (loess) curve, to diagnose potential non-linear relations,
* data ellipses, to highlight the overall trend and variability,
* point labels for potentially outlying or influential points.

I use `ggplot2` here.  It provides most of these features, except that
to label unusual points, I calculate the Mahalanobis squared distance
of all points from the grand means.

```{r lit-pers-scat}
gdf <- Guerry[, c("Literacy", "Crime_pers", "Department")]
gdf$dsq <- mahalanobis(gdf[,1:2], colMeans(gdf[,1:2]), cov(gdf[,1:2]))

ggplot(aes(x=Literacy, y=Crime_pers/1000, label=Department), data=gdf) +
  geom_point(size=2) +
  stat_ellipse(level=0.68, color="blue", size=1.2) +  
  stat_ellipse(level=0.95, color="gray", size=1, linetype=2) + 
  geom_smooth(method="lm", formula=y~x, fill="lightblue") +
  geom_smooth(method="loess", formula=y~x, color="red", se=FALSE) +
  geom_label_repel(data = gdf[gdf$dsq > 4.6,]) +
  theme_bw()
```

The flat (blue) regression line and the nearly circular data ellipses show that the correlation is nearly zero; the smoothed (red) curve indicates that there is no tendency for a nonlinear relation.

Doing the same for crimes against property:
```{r lit-prop-scat}
gdf <- Guerry[, c("Literacy", "Crime_prop", "Department")]
gdf$dsq <- mahalanobis(gdf[,1:2], colMeans(gdf[,1:2]), cov(gdf[,1:2]))

ggplot(aes(x=Literacy, y=Crime_prop/1000, label=Department), data=gdf) +
  geom_point(size=2) +
  stat_ellipse(level=0.68, color="blue", size=1.2) +  
  stat_ellipse(level=0.95, color="gray", size=1, linetype=2) + 
  geom_smooth(method="lm", formula=y~x, fill="lightblue") +
  geom_smooth(method="loess", formula=y~x, color="red", se=FALSE) +
  geom_label_repel(data = gdf[gdf$dsq > 4.6,]) +
  theme_bw()
```

### Reconnaisance plots

Reconnaisance plots attempt to give a bird's-eye overview of a multivariate data set.
For example, to see the relations among more than two variables we could turn to a
scatterplot matrix or some other display to show all pairwise bivariate relations. 


For these, my preferred package is `car` with the `scatterplotMatrix` function.
`GGally` works within the the `ggplot2` framework, but doesn't have the flexibility
I'd like.

```{r spm1, out.width="100%"}
library(car)
scatterplotMatrix(Guerry[,4:9],
                  ellipse=list(levels=0.68), 
                  smooth=FALSE)
```

Sometimes, particularly with more variables than this, we want to see a more schematic overview.  The `corrgram` package provides a variety of rendering methods to show the correlations among all pairs of variables. The panels in the
upper and lower triangles can be rendered differently.

```{r corrgram1}
library(corrgram)
corrgram(Guerry[,4:9], upper=panel.pie)
```
Or, the data in each pairwise tile can be rendered with data ellipses and
smoothed curves to show possible nonlinear relations.

```{r corrgram2}
corrgram(Guerry[,4:9], upper=panel.ellipse)
```

## Biplots

Rather than viewing the data in **data space**, a biplot shows the data in the **reduced-rank PCA space** that explains most of the variation of the observations.  This is essentially a plot of the observation scores on the first principal component overlaid with vectors representing the variables projected into PCA space.


