---
title: "Guerry data: Multivariate Analysis"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Guerry data: Multivariate Analysis}
  %\VignetteKeywords{crime, literacy, suicide, France, multivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.height = 5,
  fig.width = 5
)
```

André-Michel Guerry's (1833) _Essai sur la Statistique Morale de la France_
collected data on crimes, suicide, literacy and other "moral statistics" for various départements
in France. He provided the first real social data analysis, using graphics and maps to summarize this
multivariate dataset. One of his main goals in this ground-breaking study was to determine if the prevalence
of crime in France could be explained by other social variables.

In 1833, the scatterplot had not yet been invented; the idea of a correlation or a regression was still 50 years
in the future [@Galton:1886]. Guerry displayed his data in shaded choropleth maps and semi-graphic tables
and argued how these could be seen as implying systematic, lawful relations among moral variables.

In this analysis, we ignore the spatial context of the départements and focus on multivariate analyses of the
the data set.

## Load the data and packages

```{r load}
library(Guerry)
library(car)
library(ggplot2)
library(ggrepel)
data(Guerry)
```

## `Guerry` data set

Guerry's (1833) data consisted of six main moral variables shown in the table below. 
He wanted all of these to be recorded on aligned scales so that larger numbers consistently
reflected "morally better". Thus, four of the variables are recorded in the inverse form,
"Population per ...".


| Name        | Description |
|:------------|:------------|
|`Crime_pers` | Population per crime against persons|
|`Crime_prop` | Population per crime against property|
|`Literacy`   | Percent of military conscripts who can read and write|
|`Donations`  | Donations to the poor|
|`Infants`    | Population per illegitimate birth|
|`Suicides`   | Population per suicide|

The `Guerry` data set also contains:

* `dept` and `Department`, the French ID numbers and names for the 86 départements of metropolitan France in 1830, including Corsica.
* `Region`: a factor with main levels "N", "S", "E", "W". Corsica is coded as `NA`.
* A collection of 14 other related variables from other sources at the same time. See `?Guerry` for their precise definitions.

```{r guerry-more-vars}
names(Guerry)[-(1:9)]
```

## Guerry's questions

The main questions that concerned Guerry were whether indicators of crime could be shown to be related to factors
which might be considered to ameliorate crime. Among these, Guerry focused most on `Literacy` defined as the
number of military conscripts who could do more than mark an "X" on their enrollment form.
Other potential explanatory variables are:

: `Donations` (a measure of donations to the poor),
: `Clergy` (the rank of number of Catholic priests in active service, per population)

## Load the data and packages

```{r load}
library(Guerry)
library(car)
data(Guerry)
```



## Bivariate relationships

Let's start with plots of crime (`Crime_pers` and `Crime_prop`) in relation to `Literacy`.
A simple scatterplot is not very informative.

```{r lit-pers-scat0}
ggplot(aes(x=Literacy, y=Crime_pers/1000), data=Guerry) +
  geom_point(size=2) 

```


More useful scatterplots are annotated with additional
statistical summaries to aid interpretation:

* linear regression line,
* smoothed non-parametric (loess) curve, to diagnose potential non-linear relations,
* data ellipses, to highlight the overall trend and variability,
* point labels for potentially outlying or influential points.

I use `ggplot2` here.  It provides most of these features, except that
to label unusual points, I calculate the mahalanobis squared distance
of all points from the grand means.

```{r lit-pers-scat}
gdf <- Guerry[, c("Literacy", "Crime_pers", "Department")]
gdf$dsq <- mahalanobis(gdf[,1:2], colMeans(gdf[,1:2]), cov(gdf[,1:2]))

ggplot(aes(x=Literacy, y=Crime_pers/1000, label=Department), data=gdf) +
  geom_point(size=2) +
  stat_ellipse(level=0.68, color="blue", size=1.2) +  
  stat_ellipse(level=0.95, color="gray", size=1, linetype=2) + 
  geom_smooth(method="lm", formula=y~x, fill="lightblue") +
  geom_smooth(method="loess", formula=y~x, color="red", se=FALSE) +
  geom_label_repel(data = gdf[gdf$dsq > 4.6,]) +
  theme_bw()
```

Doing the same for crimes against property:
```{r lit-prop-scat}
gdf <- Guerry[, c("Literacy", "Crime_prop", "Department")]
gdf$dsq <- mahalanobis(gdf[,1:2], colMeans(gdf[,1:2]), cov(gdf[,1:2]))

ggplot(aes(x=Literacy, y=Crime_prop/1000, label=Department), data=gdf) +
  geom_point(size=2) +
  stat_ellipse(level=0.68, color="blue", size=1.2) +  
  stat_ellipse(level=0.95, color="gray", size=1, linetype=2) + 
  geom_smooth(method="lm", formula=y~x, fill="lightblue") +
  geom_smooth(method="loess", formula=y~x, color="red", se=FALSE) +
  geom_label_repel(data = gdf[gdf$dsq > 4.6,]) +
  theme_bw()
```





## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

