---
title: "Guerry data: Multivariate Analysis"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guerry data: Multivariate Analysis}
  %\VignetteKeywords{crime, literacy, suicide, France, multivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

André-Michel Guerry's (1833) _Essai sur la Statistique Morale de la France_
collected data on crimes, suicide, literacy and other "moral statistics" for various départements  
in France. He provided the first real social data analysis, using graphics and maps to summarize this 
multivariate dataset. One of his main goals in this ground-breaking study was to determine if the prevalence
of crime in France could be explained by other social variables.

In 1833, the scatterplot had not yet been invented; the idea of a correlation or a regression was still 50 years
in the future [@Galton:1886]. Guerry displayed his data in shaded choropleth maps and semi-graphic tables
and argued how these could be seen as implying systematic, lawful relations among moral variables.

In this analysis, we ignore the spatial context of the départements and focus on multivariate analyses of the
the data set.

## `Guerry` data set

Guerry's (1833) data consisted of six main moral variables shown in the table below. 
He wanted all of these to be recorded on aligned scales so that larger numbers consistently
reflected "morally better". Thus, four of the variables are recorded in the inverse form,
"Population per ...".


| Name        | Description |
|:------------|:------------|
|`Crime_pers` | Population per crime against persons|
|`Crime_prop` | Population per crime against property|
|`Literacy`   | Percent of military conscripts who can read and write|
|`Donations`  | Donations to the poor|
|`Infants`    | Population per illegitimate birth|
|`Suicides`   | Population per suicide|

The `Guerry` data set also contains:

* `dept` and `Department`, the French ID numbers and names for the 86 départements of metropolitan France in 1830, including Corsica.
* `Region`: a factor with main levels "N", "S", "E", "W". Corsica is coded as `NA`.
* A collection of 14 other related variables from other sources at the same time. See `?Guerry` for their precise definitions.

```{r guerry-more-vars}
names(Guerry)[-(1:9)]
```

## Guerry's questions

The main questions that concerned Guerry were whether indices of crimes could be shown to be related to factors
which might be considered to ameliorate crime. Among these, Guerry focused most on `Literacy` defined as the
number of military conscripts who could do more than mark an "X" on their enrollment form.
Other potential explanatory variables are:

: `Donations` (a measure of donations to the poor),
: `Clergy` (the rank of number of Catholic priests in active service, per population)

## Load the data and packages

```{r load}
library(Guerry)
library(car)
data(Guerry)
```


## Bivariate relationships

Let's start with plots of crime (`Crime_pers` and `Crime_prop`) in relation to `Literacy`.






Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
