### Univariate plots

```{r echo=FALSE}
data(Guerry, package = "Guerry")
library(dplyr)
library(ggplot2)
```

#### Plots of means

In his discussion, Guerry often referred to differences among the Regions of France.
One useful display to address this is a **profile** plot of the means of the moral
variables for each `Region`, typically shown as a line plot of the mean value
over the variables. Another term used for this is a **parallel coordinates** plot.

```{r sumry}
guerry_sumry <- Guerry[,c(2,4:9)] |>
  filter(!is.na(Region)) |>
  mutate(Region = factor(Region, 
                         labels =c("Central", "East", "North", "South", "West"))) |>
  group_by(Region) |>
  summarise(Crime_pers=  mean(Crime_pers),
            Crime_prop = mean(Crime_prop),
            Literacy =   mean(Literacy),
            Donations =  mean(Donations),
            Infants =    mean(Infants),
            Suicides =   mean(Suicides),
            ) |>
  ungroup() |>
  mutate_at(vars(-Region), scales::rescale)

guerry_sumry |> knitr::kable(digits = 2)
```
Even in this simple table, the regions that stand out as best (1) and worst (0)
can be seen. In particular, the Central, North and West are highest on personal crime,
while personal crime is highest in the Central region and lowest in the South.

A plot of these means by region can be produced with the `ggpcp` package.

```{r ggpcp}
library(ggpcp)
guerry_sumry |>
  pcp_select(Region, Crime_pers:Suicides) |>
  pcp_scale(method="uniminmax") |>
  pcp_arrange() |>
  ggplot(aes_pcp()) + 
  geom_pcp_axes() + 
  geom_pcp(aes(colour = Region), linewidth = 1) +
  geom_pcp_labels() +
  xlab("Variable") +
  ylab("Percent of maximum") +
  theme_bw(base_size = 14) +
  theme(legend.position = 'bottom') 
```


A more attractive form for this display shows the axes for each variable in polar coordinates,
known as a **radar chart**. Because the variables are on different scales, a first step is to
normalize them, by scaling each to a range of (0,1). The `scales::rescale()` function handles this.


Here, I use the `ggradar` package by Ricardo Bion. It is not on CRAN, so you may need to install
it to reproduce the figure here.

```{r ggradar}
#| fig.width = 8
if(!require(ggradar)) remotes::install_github("ricardo-bion/ggradar")
library(ggradar)   # Create radar charts using ggplot2

guerry_sumry |>
  ggradar(
    grid.label.size = 4, 
    axis.label.size = 3,
    axis.label.offset = 1.2,
    group.point.size = 4,
    plot.extent.x.sf = 1,
    plot.extent.y.sf = 1.2,
    fill = TRUE,
    fill.alpha = 0.2,
    legend.position = "bottom",
    legend.title = "Region",
    legend.text.size = 12,
    plot.title = "Guerry data: Means by Region")
```

