### Univariate plots

#### Plots of means

In his discussion, Guerry often referred to differences among the Regions of France.
One useful display to address this is a **profile** plot of the means of the moral
variables for each `Region`, typically shown as a line plot of the mean value
over the variables. Another term used for this is a **parallel coordinates** plot.

**TODO**: Add a parallel coordinates plot


A more attractive form for this display shows the axes for each variable in polar coordinates,
known as a **radar chart**. Because the variables are on different scales, a first step is to
normalize them, by scaling each to a range of (0,1). The `scales::rescale()` function handles this.

```{r}
guerry_sumry <- Guerry[,c(2,4:9)] |>
  filter(!is.na(Region)) |>
  group_by(Region) |>
  summarise(Crime_pers=  mean(Crime_pers),
            Crime_prop = mean(Crime_prop),
            Literacy =   mean(Literacy),
            Donations =  mean(Donations),
            Infants =    mean(Infants),
            Suicides =   mean(Suicides),
            ) |>
  ungroup() |>
  mutate_at(vars(-Region), scales::rescale)

guerry_sumry
```
Even in this simple table, the regions that stand out as best (1) and worst (0)
can be seen

Here, I use the `ggradar` package by Ricardo Bion. It is not on CRAN, so you may need to install
it to reproduce the figure here.

```{r ggradar}
#| fig.width = 7
if(!require(ggradar)) remotes::install_github("ricardo-bion/ggradar")
library(ggradar)   # Create radar charts using ggplot2

guerry_sumry |>
  ggradar(
    grid.label.size = 4, 
    axis.label.size = 4,
    group.point.size = 4,
    legend.position = "bottom",
    legend.title = "Region",
    plot.title = "Guerry data: Means by Region")
```

